SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-02-10 paper/paper.tex
paper updates.

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered by QuickCheck.  Implementations of \emph{shrink}
often require significant boilerplate and usually just generate smaller tests
than the discovered counterexample.  Therefore, shrinking can be inefficient and
poor at discovering new, smaller counterexamples.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples and make it easier for the programmer to uncover multiple
counterexamples at once.  SmartCheck has been implemented for Haskell and is
freely available.
\end{abstract}

@@

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
Indeed, because of the regular structure of \ttp{shrink} implementations, the
principal motivation for one of Haskell's first generics package was for
providing generic definitions for \ttp{shrink}~\cite{syb}.  However, simple
structural recursion may either miss smaller counterexamples, be too inefficient
to be practical, or both.\footnote{For a recent real-world case, see the user
  question posed on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell.  We strive to use idiomatic
  Haskell, and only rely on the language extensions of existential type
  quantification and scoped type variables.~\cite{haskell98}.}  Data type
\ttp{T} is a product type containing four lists of \ttp{Int16}s and a single
\ttp{Word8}.  For our purposes, it is not important what \ttp{T} models, but
for concreteness, suppose it models the values in four linked lists in an
operating system, together with an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$.  At first glance, the property seems reasonable, but we
have forgotten about underflows.  For example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
fails \ttp{prop} (\ttp{==>} is implication from the QuickCheck library).
Unfortunately, a typical counterexample returned by QuickCheck is not quite so
simple.  With standard settings, it returns a value containing just under 70
\ttp{N} constructors, and sometimes values contain over 100 constructors.  Thus,
it pays to define \ttp{shrink}!

A typical definition of \ttp{shrink} for a data type like \ttp{T} is to take the
cross-product of shrunk values over the arguments to the constructor \ttp{T}.
This can be expressed using Haskell's list-comprehension notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
The result of (\ttp{shrink t}) with this definition is null if any list
contained in \ttp{t} is null.  Still, for typical counterexamples returned by
QuickCheck that do not contain empty lists, together with QuickCheck's default
\ttp{Arbitrary} instances, the length of potential counterexamples returned by
\ttp{shrink} can easily exceed $10^{10}$, which is an intractable number of
tests for QuickCheck to analyze.  The time complexity is compounded by the
shrink stage being recursive, so when a new smaller counterexample is found, it
is also shrunk.  In practice, QuickCheck appears to loop indefinitely.

It is not always straightforward to determine how to make a more efficient
implementation for \ttp{shrink}.  Shrinking presents a programmer's dilemma:
shrinking is used to better understand a counterexample to understand why a
property failed, but to implement an efficient shrink function in general
requires knowing how to generate counterexamples to the property.

Typically, a programmer tries a brute-force approach to reduce the number of
potential counterexamples; for example, she may try to truncate lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, we might redefine shrink as follows:
%
%\begin{figure}[ht]
\begin{code}
len = 10

shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take len (shrink x)
\end{code}
%%   \caption{Truncated shrinking function for data type \ttp{B}.}
%%   \label{lst:newshrink}
%% \end{figure}

%% \end{lstlisting}
%% \end{minipage}
%
\noindent
let us call this version ``truncated shrink''.  Truncating the controls the
blowup of the input-space; the number of possible new values is $10^5$ in this
case.  While truncation controls the blowup, the downside is that potentially
smaller counterexamples may be omitted.

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
 & QC (none) & QC (10) & QC (20) & SmartCheck \\
\hline \hline
Max.  & 0.15s & 21.51s & 125.37s & 1.96s\\
\hline
Mean  & 0.07s & 1.21s & 3.80s & 0.30s\\
\hline
Median & 0.07s & 0.48s & 0.52s & 0.24s\\
\hline
Mean size & 70 & 34 & 32 & 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 100 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{table:results}
\end{table}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider
Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is not even feasible for an example like this).  In the
first three columns, the results of using QuickCheck to generate counterexamples
are shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of truncated \ttp{shrink} that limit the number of
smaller values to 10 and 20 elements, respectively.  (With the original
definition for shrink from Figure~\ref{fig:initial}, some counterexamples cause
QuickCheck to not return during the shrinking phase after hours of computation,
so we omit these.)

In each row, we show the maximum, mean, and median number of seconds required by
QuickCheck to generate a counterexample over 100 runs.  In the last row, we show
the mean size of the final value returned, measured by the taking the cumulative
length of the lists of the fields of \ttp{B}.  Notice that while typically
counterexamples are returned in a fraction of a second, there are outliers
taking 21 seconds and 125 seconds for the two tests.  Both tests roughly cut the
size of the counterexample in half; increasing the number of potential
counterexamples can result in a significant performance penalty for this example
without much benefit.

In the last column are the results for SmartCheck.  Not only does it execute
faster than QuickCheck with shrinking, it has significantly smaller outliers.
The most striking difference, however, is the size of the counterexample
returned, reducing on average the size of a counterexample found by QuickCheck
without shrinking by a factor of 14.

The purpose of this paper is to explain first how to efficiently and generically
generate small counterexamples for algebraic data types and second how to
generalize the counterexamples.

\todo{check paper against this section again, especially generalization sec.}

@@

\noindent
In Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from Xmonad.

@@

\todo{make clear that preconditions shouldn't be considered logical
  preconditions}

@@

To present the algorithms in the following sections, we provide some definitions
here.  Our presentation simplifies the implementation
(Section~\ref{sec:implementation}) somewhat to improve the presentation.
Consider the following types and type class.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with indexes in a breadth-first order:
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType L          = False
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.

In our implementation, \ttp{constr} and \ttp{constrs} depends on GHC
Generics~\cite{generics}, which we describe in Section~\ref{sec:implementation}.
For simplicity, we omit here Generics-specific super-class constraints on the
\ttp{SubTypes} class.

@@

newVals :: SubVal -> Size -> Int -> IO [SubVal]
newVals (SubVal a) sz tries =
  replicateM tries s
  where
  s  = liftM SubVal (sizedArbitrary sz a)

reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v (getSize v) scMaxReduce
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index
\ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the algorithm
terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successfully larger counterexamples when generating new values with which to
replace a sub-value.

\todo{explain getSize}

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, let us assume one is generated locally).  The constant
\ttp{scMaxReduce} is the maximum number of tries to discover a new
counterexample by replacing \ttp{v} in \ttp{cex} and testing it.  The result of
\ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  The property may be a
conditional, in which case the value must pass the precondition as well as the
consequent for \ttp{pass} to return \ttp{True}.  If no failure is found, we move
to the next sub-value of \ttp{cex} and continue.  However, if a new smaller
counterexample \ttp{cex'} is found, we start a new breadth-first traversal of
\ttp{cex'}, attempting to shrink it further.  The algorithm is guaranteed to
terminate: intuitively, the measure for the function is that either the index
increases or the size of the counterexample being evaluated decreases.  The
complexity of the algorithm is $\Omega(n^2)$, where $n$ is the number of
constructors in the counterexample.  The analysis assumes generating new
sub-values and testing them is done in constant time.

@@

\begin{figure}
\begin{code}
reduce0 :: forall a . SubTypes a
        => a -> (a -> Property) -> IO a
reduce0 cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> case testHole v of
                   Just v' -> reduce0 v' prop
                   Nothing -> test' v idx

  test' v idx = do
    vs <- newVals v (getSize v) scMaxReduce
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

  testHole (SubVal a) =
    case cast a :: Maybe a of
      Nothing -> Nothing
      Just a' ->
        if pass prop a' then Nothing
          else Just a'
\end{code}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

Generalizing the scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.  In
Figure~\ref{fig:reduce0}, we show an updated reduction algorithm, \ttp{reduce0}
that implements this optimization.  The function \ttp{testHole} tests the
current sub-value and if it fails the property, the function tests the sub-value
directly.

@@

extrapolate :: SubTypes a
            => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
    | otherwise
    = case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scMaxExtrap
          extrapolate' (idx+1)
            (if allFail cex idx vs prop
               then idx:idxs else idxs)

allFail :: SubTypes a
        => a -> Idx -> [SubVal]
        -> (a -> Property) -> Bool
allFail cex idx vs prop =
  length res >= scMinExtrap && and res
  where
  res  = mapMaybe go vs
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

@@

\begin{figure}
  \begin{code}
matchesShapes :: SubTypes a => a -> [(a,([Idx],[Idx]))] -> Bool
matchesShapes d = any (matchesShape d)

matchesShape :: SubTypes a => a -> (a, ([Idx],[Idx])) -> Bool
matchesShape a (b, (idxVals, idxConstrs))
  | constr a /= constr b = False
  | Just a' <- aRepl         = let x = subForest (subVals a') in
                               let y = subForest (subVals b) in
                               all foldEqConstrs (zip x y)
  | otherwise                = False

  where
  foldEqConstrs :: (Tree SubVal, Tree SubVal) -> Bool
  foldEqConstrs (Node (SubVal l0) sts0, Node (SubVal l1) sts1)
    | constr l0 == constr l1 = next
    | otherwise              = False
    where next = all foldEqConstrs (zip sts0 sts1)

  updateA idx d = fmap (replace d idx) (index b idx)
  aRepl = foldl go (Just a) (idxVals ++ idxConstrs)
    where go ma idx = maybe Nothing (updateA idx) ma
  \end{code}
  \caption{Shape matching algorithm.}
  \label{fig:matches}
\end{figure}

@@

sumTest :: SubTypes a
          => a -> (a -> Property) -> [Idx] -> IO [Idx]
sumTest cex prop exIdxs = sumTest' 1 []
  where
  sumTest' idx idxs
    | subTrees cex idx (exIdxs ++ idxs)
    = sumTest' (idx+1) idxs
    | otherwise
    = case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scConstrMax
          sumTest' (idx+1)
            (if constrFail cex idx vs prop
                  (subConstr v) (subConstrs v)
               then idx:idxs else idxs)

>>>>

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered by QuickCheck.  Implementations of shrink often
require significant boilerplate and usually just generate smaller tests than the
discovered counterexample.  Therefore, shrinking can be inefficient and poor at
discovering new, smaller counterexamples.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples and make it easier for the programmer to uncover multiple
counterexamples at once.  SmartCheck has been implemented for Haskell and is
freely available.
\end{abstract}

@@

In industrial uses, shrinking is essential.  Hughes, the inventor of QuickCheck
and the CEO of Quiviq, a company that commercializes the technology, has noted
that ``without it [shrinking], randomly generated failing cases would often be
so large as to be almost useless.''~\cite{qcjh}.  Hughes~\emph{et al.} give an
extended example in which shrinking is essential in debugging telecom
software~\cite{telecom}.

Typically, effective shrinking on complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.
Shrinking presents a programmer's dilemma: shrinking is used to better
understand a counterexample to understand why a property failed, but to
implement an efficient shrink function in general requires knowing how to
generate counterexamples to the property.

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in Haskell.  We strive to
  use idiomatic Haskell, and only rely on the language extensions of existential
  type quantification and scoped type variables.~\cite{haskell98}.}  Data type
\ttp{T} is a product type containing four lists of \ttp{Int16}s and a single
\ttp{Word8}.  For our purposes, it is not important what \ttp{T} models, but for
concreteness, suppose it models the values in four linked lists in an operating
system, together with an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
fails \ttp{prop} (\ttp{==>} is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck is not quite so simple.  With standard settings, the median
counterexample discovered contains around 70 \ttp{N} constructors, and
counterexamples can contain over 100 constructors.  Thus, it pays to define
\ttp{shrink}!

We might first naively try to shrink counterexamples for a data type like \ttp{T}
by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, then \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to loop for hours, consuming
ever more memory, without returning a result.

A programmer might try to control the complexity by truncating lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, we might redefine shrink as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space; the maximum number of possible new values is $10^5$ in this case.
While truncation controls the blowup, the downside is that potentially smaller
counterexamples may be omitted.

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  let xs = shrink (w, i0, i1, i2, i3) in
  let go (w', i0', i1', i2', i3') =
    T w' i0' i1' i2' i3' in
  map go xs
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant, and the approach is generalized
to larger tuples.  Still, with this definition, in some cases, shrinking takes
over 20 seconds and returns a result containing over 60
constructors.\footnote{Results are with a GHC-compiled program, using \ttp{-O2},
  on a 4-core 2.7GHz Intel~i7 processor.}

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
         & QC (none)  & QC (trunc) & QC (tuple) & SmartCheck\\
\hline \hline
Maximum  & 0.40s, 124 & 139.15s, 90  & 97.53s, 53 & 0.s, 13\\
\hline
Mean     & 0.08s, 69  & 0.73s, 35  & 2.26s, 13  & 0.07s, 6\\
\hline
Median   & 0.07s, 68  & 0.18s, 32  & 1.90s, 13  & 0.07s, 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 1000 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{table:results}
\end{table}
\todo{table generated with InitEx2 on Feb. 3 --- do a final test with 1000}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider
Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as noted, it is not even feasible for the example).  We show the maximum,
mean, and median amount of time required to generate a counterexample and to
shrink it (and print it to standard out) as well as the corresponding number of
constructors in the final result.

In the first three columns, the results of using QuickCheck to generate
counterexamples are shown.  The first column contains the results of using
QuickCheck without shrinking to provide a baseline.  The next column shows the
result of using QuickCheck with truncating shrink, which each list is truncated
to 10 elements.  Next, we show the results of the ``tuple shrink'' instance.  In
the last column are the results for SmartCheck, which includes the time required
by QuickCheck to generate the original counterexample.  Not only does SmartCheck
execute faster than the other strategies, it has smaller outliers.  In addition,
the size of the counterexamples are significantly smaller.  And we stress that
SmartCheck achieves these results \emph{generically}.

\todo{change max to std dev, look for max info above.}

@@

\noindent
In Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from Xmonad, and related work and
concluding remarks are made in Sections~\ref{sec:related}
and~\ref{sec:conclusions}, respectively.

@@

@@

To present the algorithms in the following sections, we provide some definitions
in the form of methods of a type class.  SmartCheck requires instances for data
being analyzed.  In Section~\ref{sec:implementation}, we describe how instances
for the type class are derived automatically.  Consider the following types and
type class.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data T = L | B T T
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with indexes in a breadth-first order:
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType L          = False

subVals (B (B L\sub{0} L\sub{1}) L\sub{2}) =
  Tree (B L\sub{0} L\sub{1}) [Tree L\sub{0} [], Tree L\sub{1} []]
       (Tree L\sub{1} [])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.

In our implementation, \ttp{constr} and \ttp{constrs} depends on GHC
Generics~\cite{generics}, which we describe in Section~\ref{sec:implementation}.
For simplicity, we omit here Generics-specific super-class constraints on the
\ttp{SubTypes} class.  Moreover, our presentation simplifies the implementation
(Section~\ref{sec:implementation}) somewhat to improve the presentation.

@@

newVals :: Size -> Int -> SubVal -> IO [SubVal]
newVals sz tries (SubVal a) =
  replicateM tries s where
  s  = liftM SubVal (sizedArbitrary sz a)

reduce :: SubTypes a
  => ScArgs -> (a -> Property) -> a -> IO a
reduce args prop cex = reduce' 1
  where
  reduce' idx
    | Just v <- index cex idx
    = do vs <- newVals (getSize v)
                 (scMaxReduce args) v
         maybe (reduce' (idx+1)) (reduce args prop)
               (test cex idx vs prop)
    | otherwise = return cex

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes arguments to customize the algorithm's behavior, a
counterexample \ttp{cex} and the property \ttp{prop}.  At the beginning of
execution, we assume that a counterexample has been obtained somehow; our
implementation relies on QuickCheck as a ``front-end'' to generate a
counterexample that the algorithm automatically shrinks.  The reduction begins
at the first proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index
\ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the algorithm
terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successfully larger counterexamples when generating new values with which to
replace a sub-value.

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, let us assume one is generated locally).  The constant
\ttp{scMaxReduce} is the maximum number of tries to discover a new
counterexample by replacing \ttp{v} in \ttp{cex} and testing it.  The result of
\ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  The property may be a
conditional, in which case the value must pass the precondition as well as the
consequent for \ttp{pass} to return \ttp{True}.  If no failure is found, we move
to the next sub-value of \ttp{cex} and continue.  However, if a new smaller
counterexample \ttp{cex'} is found, we start a new breadth-first traversal of
\ttp{cex'}, attempting to shrink it further.  The algorithm is guaranteed to
terminate: intuitively, the measure for the function is that either the index
increases or the size of the counterexample being evaluated decreases.  The
complexity of the algorithm is $\mathcal{O}(n^2)$, where $n$ is the number of
constructors in the counterexample.  The analysis assumes generating new
sub-values and testing them is done in constant time.

@@

\begin{figure}
\begin{code}
reduceOpt :: forall a . SubTypes a
  => ScArgs -> (a -> Property) -> a -> IO a
reduceOpt args prop cex = reduce' 1
  where
  reduce' idx
    | Just v <- index cex idx
    = maybe (test' v idx) (reduceOpt args prop)
            (testHole v)
    | otherwise = return cex

  test' v idx = do
    vs <- newVals (getSize v) (scMaxReduce args) v
    maybe (reduce' (idx+1)) (reduceOpt args prop)
          (test cex idx vs prop)

  testHole (SubVal a) = do
    a' <- cast a :: Maybe a
    if pass prop a' then Nothing else Just a'
\end{code}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

Generalizing the scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.  In
Figure~\ref{fig:reduce0}, we show an updated reduction algorithm, \ttp{reduceOpt}
that implements this optimization.  The function \ttp{testHole} tests the
current sub-value and if it fails the property, the function tests the sub-value
directly.

@@

extrapolate :: SubTypes a
  => ScArgs -> a -> (a -> Property) -> IO [Idx]
extrapolate args cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
    | Just v <- index cex idx = mkNewVals v
    | otherwise = return idxs
    where
    mkNewVals v = do
      vs <- newVals (scMaxSize args)
                    (scMaxExtrap args) v
      extrapolate' (idx+1)
        (if allFail args cex idx vs prop
           then idx:idxs else idxs)

allFail :: SubTypes a => ScArgs -> a -> Idx
  -> [SubVal] -> (a -> Property) -> Bool
allFail args cex idx vs prop =
  length res >= scMinExtrap args && and res
  where
  res  = mapMaybe go vs
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

@@

@@

sumTest :: SubTypes a => ScArgs -> a
  -> (a -> Property) -> [Idx] -> IO [Idx]
sumTest args cex prop exIdxs = sumTest' 1 []
  where
  sumTest' idx idxs
    | subTrees cex idx (exIdxs ++ idxs)
    = sumTest' (idx+1) idxs
    | Just v <- index cex idx = fromSumTest v
    | otherwise = return idxs
    where
    fromSumTest v = do
      vs <- newVals (scMaxSize args)
              (scConstrMax args) v
      sumTest' (idx+1)
        (if constrFail cex idx vs prop
           (subConstr v) (subConstrs v)
           then idx:idxs else idxs)
